Semi-Parametric Manifold Clustering

### 2021-11-10 curve-fitting-methods/curve-fitting-scratch.Rmd
updated notes

```{r}
set.seed(314159)
n <- 200
t. <- runif(n)
X <- param.curve.quadr.2(t., 0, 0, 1, 0, 2, -2)

out <- estimate.quadr.curve.2(X)
thetahat <- out$theta
t.hat <- sort(out$t)
Xhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

@@

One problem with this method is the parameterization of the curve is not unique.

@@

```{r, cache = TRUE}
out1 <- estimate.quadr.curve.2(Y,
                              init.params = c(0, 0, 1, 0, 2, -2),
                              intercept = TRUE,
                              maxit = 200)
thetahat <- out1$theta
t.hat <- sort(out1$t)
Yhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

@@

@@

1. Assign an initial clustering (e.g., via spectral clustering).
2. Estimate the curve for each cluster.
3. Reassign the clusters by proximity to each curve.
4. Repeat 2 and 3 until convergence.

We again limit these to be quadratic functions in $\mathbb{R}^2$. 
Here, $K = 2$ and $n_1 = n_2 = 256$.

```{r}
f1 <- function(t) {
  x1 <- t / sqrt(.5 + sqrt(5) / 2)
  x2 <- x1 ** 2
  return(cbind(x1, x2))
}

f2 <- function(t) {
  x2 <- t / sqrt(.5 + sqrt(5) / 2)
  x1 <- x2 ** 2
  return(cbind(x1, x2))
}

n1 <- n2 <- 2 ** 8
z <- c(rep(1, n1), rep(2, n2))
t1 <- sort(runif(n1))
t2 <- sort(runif(n2))
X1 <- f1(t1)
X2 <- f2(t2)
X <- rbind(X1, X2)

P <- X %*% t(X)
Xhat <- embedding(P, 2, 0)

manifold.out <- manifold.clustering.quadr.2(P, 2, 
                                            intercept = FALSE, 
                                            method = 'polynomial')
zhat.manifold <- manifold.out$z
if (mean(z == zhat.manifold) < .5) {
  zhat.manifold <- 2 - zhat.manifold
}
t. <- seq(0, 1, .01)
theta1 <- manifold.out$theta[[1]]
theta2 <- manifold.out$theta[[2]]
Y1 <- param.curve.quadr.2(t., 
                          theta1[1], theta1[2], theta1[3], 
                          theta1[4], theta1[5], theta1[6])
Y2 <- param.curve.quadr.2(t.,
                          theta2[1], theta2[2], theta2[3],
                          theta2[4], theta2[5], theta2[6])
Y <- rbind(Y1, Y2)
z. <- c(rep(1, length(t.)),
        rep(2, length(t.)))

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw() + 
  # geom_line(aes(x = Y[, 1], y = Y[, 2],
  #               groups = factor(z.))) + 
  xlim(-1, 0) + ylim(-.5, .5) + 
  coord_fixed()
```

Next, we draw $A \sim \RDPG(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
The community detection error in this example is 30\%.
In the following plot, the points are colored according to their true labels.

```{r cache = TRUE, fig.width = 3, fig.height = 3}
set.seed(12345)

A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)

manifold.out <- manifold.clustering.quadr.2(A, 2, 
                                            intercept = FALSE,
                                            method = 'polynomial')
zhat.manifold <- manifold.out$z
# table(z, zhat.manifold)

# plot(Xhat, col = z, asp = 1)
t. <- seq(-1, 1, .01)
theta1 <- manifold.out$theta[[1]]
theta2 <- manifold.out$theta[[2]]
Y1 <- param.curve.quadr.2(t., 
                          theta1[1], theta1[2], theta1[3], 
                          theta1[4], theta1[5], theta1[6])
Y2 <- param.curve.quadr.2(t.,
                          theta2[1], theta2[2], theta2[3],
                          theta2[4], theta2[5], theta2[6])
Y <- rbind(Y1, Y2)
z. <- c(rep(1, length(t.)),
        rep(2, length(t.)))
# lines(Y1)
# lines(Y2, col = 2)

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw() + 
  geom_path(aes(x = Y[, 1], y = Y[, 2],
                groups = factor(z.))) + 
  xlim(-.25, 1.25) + ylim(-.75, .75) + 
  coord_fixed()
```

A possibly more robust modification to this is to use Bezier curves for $g$. 
This is the same functional form as the polynomial curves we used before, but with different basis functions. 
For the following examples, we consider the quadratic Bezier curve, which has the following basis form:

$$
g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
$$

Where as before, $g : [0, 1] \mapsto \mathbb{R}^d$ and $t \in [0, 1]$, and each $p_r \in \mathbb{R}^d$. Thus if we can fit each $p_r$, then the procedure is the same as before.

Common methods for Bezier curve fitting do not require specification of $t_1, ..., t_n$. However, an ordering is required. The curve fitting method is then modified as follows:

1. Initialize $\hat{\theta}^{(0)}$ (e.g., randomly). 
2. Estimate each $\hat{t}_i^{(s)}$ by minimizing $L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2$.
3. Reorder $X$ according to $t_1, ..., t_n$.
4. Estimate $\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}$ for the reordered $X$ using Bezier curve fitting.
5. Compute $\hat{\theta}^{(s+1)}$ from $\{\hat{p}_r^{(s)}\}$.
6. Repeat steps 2-5 until convergence.

@@

We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$. $n_1 = n_2 = 128$ points are drawn uniformly from each.

```{r}
n1 <- 2 ** 7
n2 <- n1
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))

@@

ggplot() + 
  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z))) + 
  coord_fixed() + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw()
```

@@

```{r}
P <- X %*% t(X)
diag(P) <- 0
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
if (mean(Xhat[, 1]) < 0) Xhat[, 1] <- -Xhat[, 1]

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
  coord_fixed() + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw()
```

@@

```{r cache = TRUE}
set.seed(314159265)
manifold.out <- manifold.clustering.quadr.2(A, 2, 
                                            # initialization = z,
                                            # initialization = 'spectral',
                                            initialization = 'random', 
                                            method = 'bezier')
zhat.manifold <- manifold.out$z

@@

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(zhat.manifold))) + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw() + 
  geom_path(aes(x = Y[, 1], y = Y[, 2],
                groups = factor(z.))) + 
  xlim(-.25, 1.25) + ylim(-.75, .75) + 
  coord_fixed()
```

@@

>>>>

```{r}
set.seed(314)
n <- 200
t. <- runif(n)
X <- param.curve.quadr.2(t., 0, 0, 1, 0, 2, -2)

out <- estimate.quadr.curve.2(X, maxit = 100)
thetahat <- out$theta
t.hat <- sort(out$t)
Xhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

@@

Note: the parameterization of the curve is not unique.

@@

```{r, cache = TRUE}
out1 <- estimate.quadr.curve.2(Y,
                              intercept = TRUE,
                              maxit = 200)
thetahat <- out1$theta
t.hat <- sort(out1$t)
Yhat <- param.curve.quadr.2(t.hat, 
                            thetahat[1], thetahat[2], thetahat[3],
                            thetahat[4], thetahat[5], thetahat[6])

@@

A modification to this that is possibly more robust is to use Bezier curves for $g$. 
This is the same functional form as the polynomial curves used before, but with orthogonal bases:

$$g(t | p) = \sum_{r = 0}^{R} p_r \binom{R}{r} (1 - t)^{R - r} t^r$$

where $R$ is the order of the Bezier curve, each $p_s$ is a vector of length $d$, and, as before, $g : [0, 1] \mapsto \mathbb{R}^d$. 
Thus, if we fit each $p_r$, then the procedure is the same as before.

The least squares estimate for $p \in \mathbb{R}^{R \times d}$ is

$$\hat{p} = (T^\top T)^{-1} T^\top X$$

where $T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}$ and each $t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
The same procedure for estimating $t_1, ..., t_n$ can be applied here.

The parameterization for a given curve is not unique. 
In particular, the above procedure will not necessarily provide $t_1, ..., t_n \in [0, 1]$. 
One possible remedy for this is to, after estimating the $t_i$'s, normalize them to the unit interval. If we assume $t_1, ..., t_n \iid Uniform(a, b)$, then the UMVUE are

$$\hat{a} = \frac{n t_{(1)} - t_{(n)}}{n - 1},
\hat{b} = \frac{n t_{(n)} - t_{(1)}}{n - 1}$$

which yields the normalization transformation $t \leftarrow (t - \hat{a}) / (\hat{b} - \hat{a})$.

Alternatively, we can force the $t_i$'s to be approxiately uniform on the unit interval by the transformation $t \leftarrow \hat{F}(t)$, where $\hat{F}$ is the empirical CDF of $t_1, ..., t_n$.

For initialization, we can use a one-dimensional Isomap embedding to estimate the $t_i$'s and use that to estimate $\hat{p}$.
Experiments suggest that if the data are well-behaved (i.e., it looks like the curve we are trying to fit), this results in much faster convergence.

@@

1. Assign an initial clustering (e.g., via spectral clustering).
2. Estimate the curve for each cluster (using the same curve-fitting procedure as before).
3. Reassign the clusters by proximity to each curve.
4. Repeat 2 and 3 until convergence.

@@

We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$. $n_1 = n_2 = 256$ points are drawn uniformly from each.

```{r}
set.seed(314159265)
n1 <- 2 ** 8
n2 <- n1
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))

@@

ggplot() + 
  geom_point(aes(x = X[, 1], y = X[, 2], colour = factor(z))) + 
  coord_fixed() + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw()

P <- X %*% t(X)
diag(P) <- 0
A <- draw.graph(P)
Xhat <- embedding(A, 2, 0)
# if (mean(Xhat[, 1]) < 0) Xhat[, 1] <- -Xhat[, 1]
```

@@

```{r, fig.height = 3}
ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(z))) + 
  coord_fixed() + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw()
```

@@

```{r cache = TRUE, eval = TRUE}
set.seed(314159)
manifold.out <- manifold.clustering.quadr.2(A, 2, 
                                            # initialization = z,
                                            # initialization = 'spectral',
                                            initialization = 'random',
                                            method = 'bezier')

zhat.manifold <- manifold.out$z

@@

ggplot() + 
  geom_point(aes(x = Xhat[, 1], y = Xhat[, 2], colour = factor(zhat.manifold))) + 
  labs(x = expression(x[1]), y = expression(x[2]), colour = NULL) + 
  theme_bw() + 
  geom_path(aes(x = Y[, 1], y = Y[, 2],
                group = factor(z.))) +
  xlim(min(Xhat[, 1]), max(Xhat[, 1])) + ylim(min(Xhat[, 2]), max(Xhat[, 2])) + 
  coord_fixed()
```

@@

## Model-Based Clustering

The CLT property of the ASE may allow us to approximate $W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)$.
One possibly reasonable assumption for $\Sigma_k$ is $\Sigma_k = \Sigma_k(t) =  \diag(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))$. A plot of the squared errors of cluster 1 in the previous example reveals the following:

```{r, fig.height = 2, fig.width = 8}
theta1 <- manifold.out$theta[[1]]
theta2 <- manifold.out$theta[[2]]
t. <- manifold.out$t
X1 <- param.curve.quadr.2(t.[[1]], 
                          theta1[1], theta1[2], theta1[3],
                          theta1[4], theta1[5], theta1[6])
X2 <- param.curve.quadr.2(t.[[2]], 
                          theta2[1], theta2[2], theta2[3],
                          theta2[4], theta2[5], theta2[6])
# plot(Xhat[zhat.manifold == 1, 1] - X1[, 1])
# plot(Xhat[zhat.manifold == 1, 2] - X1[, 2])

p1 <- ggplot() + 
  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 1] - X1[, 1]) ^ 2)) + 
  theme_bw() + 
  labs(x = 't', y = expression((x[1] - hat(x)[1])^2))

p2 <- ggplot() + 
  geom_point(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
  stat_smooth(aes(x = t.[[1]], y = (Xhat[zhat.manifold == 1, 2] - X1[, 2]) ^ 2)) + 
  theme_bw() + 
  labs(x = 't', y = expression((x[2] - hat(x)[2])^2))

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

Fitting some curve to each $\sigma_{kr}^2(t)$ allows us to compute estimated probabilities of each point belonging to each cluster/manifold: $\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))$. 
Then we can use weighted least squares to estimate the parameters of $g$:

$$\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X$$

where $Q_k = \diag(q_{1k}, ..., q_{nk})$.

### 2021-11-10 curve-fitting-methods/curve-fitting-scratch.tex
updated notes

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Assign an initial clustering (e.g., via spectral clustering).
\item
  Estimate the curve for each cluster.
\item
  Reassign the clusters by proximity to each curve.
\item
  Repeat 2 and 3 until convergence.
\end{enumerate}

@@

We again limit these to be quadratic functions in $\mathbb{R}^2$. 
Here, $K = 2$ and $n_1 = n_2 = 256$.

\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-6-1} \end{center}

Next, we draw $A \sim \mathrm{RDPG}(X)$ from the above example and apply the clustering and curve fitting to the ASE of $A$.
The community detection error in this example is 30\%.
In the following plot, the points are colored according to their true labels.

@@

@@

A possibly more robust modification to this is to use Bezier curves for
\(g\). This is the same functional form as the polynomial curves we used
before, but with different basis functions. For the following examples,
we consider the quadratic Bezier curve, which has the following basis
form:

\[
g(t) = (1 - t)^2 p_0 + 2 t (1 - t) p_1 + t^2 p_2
\]

Where as before, \(g : [0, 1] \mapsto \mathbb{R}^d\) and
\(t \in [0, 1]\), and each \(p_r \in \mathbb{R}^d\). Thus if we can fit
each \(p_r\), then the procedure is the same as before.

Common methods for Bezier curve fitting do not require specification of
\(t_1, ..., t_n\). However, an ordering is required. The curve fitting
method is then modified as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Initialize \(\hat{\theta}^{(0)}\) (e.g., randomly).
\item
  Estimate each \(\hat{t}_i^{(s)}\) by minimizing
  \(L(t_i, \hat{\theta}^{(s)} | x_i) = L_i = \|x_i - g(t_i | \hat{\theta}^{(s)})\|^2\).
\item
  Reorder \(X\) according to \(t_1, ..., t_n\).
\item
  Estimate \(\hat{p}_0^{(s)}, \hat{p}_1^{(s)}, \hat{p}_2^{(s)}\) for the
  reordered \(X\) using Bezier curve fitting.
\item
  Compute \(\hat{\theta}^{(s+1)}\) from \(\{\hat{p}_r^{(s)}\}\).
\item
  Repeat steps 2-5 until convergence.
\end{enumerate}

We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$. $n_1 = n_2 = 128$ points are drawn uniformly from each.

@@

\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-11-1} \end{center}

>>>>

A modification to this that is possibly more robust is to use Bezier
curves for \(g\). This is the same functional form as the polynomial
curves used before, but with orthogonal bases:

\[g(t | p) = \sum_{r = 0}^{R} p_r \binom{R}{r} (1 - t)^{R - r} t^r\]

where \(R\) is the order of the Bezier curve, each \(p_s\) is a vector
of length \(d\), and, as before, \(g : [0, 1] \mapsto \mathbb{R}^d\).
Thus, if we fit each \(p_r\), then the procedure is the same as before.

The least squares estimate for \(p \in \mathbb{R}^{R \times d}\) is

\[\hat{p} = (T^\top T)^{-1} T^\top X\]

where \(T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}\)
and each \(t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
\(X \in \mathbb{R}^{n \times d}\). The same procedure for estimating
\(t_1, ..., t_n\) can be applied here.

The parameterization for a given curve is not unique. In particular, the
above procedure will not necessarily provide
\(t_1, ..., t_n \in [0, 1]\). One possible remedy for this is to, after
estimating the \(t_i\)'s, normalize them to the unit interval. If we
assume \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}Uniform(a, b)\),
then the UMVUE are

\[\hat{a} = \frac{n t_{(1)} - t_{(n)}}{n - 1},
\hat{b} = \frac{n t_{(n)} - t_{(1)}}{n - 1}\]

which yields the normalization transformation
\(t \leftarrow (t - \hat{a}) / (\hat{b} - \hat{a})\).

Alternatively, we can force the \(t_i\)'s to be approxiately uniform on
the unit interval by the transformation \(t \leftarrow \hat{F}(t)\),
where \(\hat{F}\) is the empirical CDF of \(t_1, ..., t_n\).

For initialization, we can use a one-dimensional Isomap embedding to
estimate the \(t_i\)'s and use that to estimate \(\hat{p}\). Experiments
suggest that if the data are well-behaved (i.e., it looks like the curve
we are trying to fit), this results in much faster convergence.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Assign an initial clustering (e.g., via spectral clustering).
\item
  Estimate the curve for each cluster (using the same curve-fitting
  procedure as before).
\item
  Reassign the clusters by proximity to each curve.
\item
  Repeat 2 and 3 until convergence.
\end{enumerate}

@@

We have two intersecting curves, $g_1(t) = \begin{bmatrix} t^2 & 2 t (1 - t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1 - t) & (1 - t) ^ 2 \end{bmatrix}^\top$. $n_1 = n_2 = 256$ points are drawn uniformly from each.

@@

\begin{center}\includegraphics{curve-fitting-scratch_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

\hypertarget{model-based-clustering}{%
\subsection{Model-Based Clustering}\label{model-based-clustering}}

The CLT property of the ASE may allow us to approximate
\(W x - \hat{x} \sim \mathcal{N}(f_k(t), \Sigma_k)\). One possibly
reasonable assumption for \(\Sigma_k\) is
\(\Sigma_k = \Sigma_k(t) = \mathrm{diag}(\sigma_{k1}^2(t), ..., \sigma_{kd}^2(t))\).
A plot of the squared errors of cluster 1 in the previous example
reveals the following:

@@

Fitting some curve to each \(\sigma_{kr}^2(t)\) allows us to compute
estimated probabilities of each point belonging to each
cluster/manifold:
\(\hat{q}_{ik} = \mathcal{N}(x_i \mid g_k(\hat{t}_i), \Sigma_k(\hat{t}_i))\).
Then we can use weighted least squares to estimate the parameters of
\(g\):

\[\hat{p}_k = (T^\top Q_k T)^{-1} T^\top Q_k X\]

where \(Q_k = \mathrm{diag}(q_{1k}, ..., q_{nk})\).
